Cryptobenthic reef assemblages are more distinct across a 90 m depth gradient than 2,500 km of shallow marine habitat in the Hawaiian Archipelago

### 2025-03-22 manuscript.Rmd
rearrangements and updates for sci reports

The Hawaiian Archipelago, stretching across ~2,500 km in the North Pacific Ocean, provides a natural laboratory to examine drivers of metazoan community structure. This chain of islands comprises two regions, the inhabited high islands of the main Hawaiian Islands (MHI) and the uninhabited low islands, atolls, and banks of the Northwestern Hawaiian Islands (NWHI) which reside within one of the world's largest marine protected areas---the Papahānaumokuākea Marine National Monument. The island chain is subject to considerable anthropogenic and physical gradients driving differences in temperature, coral cover, human population, and anthropogenic impact, among others, all of which influence coral reef communities [@selkoeMapHumanImpacts2009]. Nearshore drop-offs along the island chain also allow access to deeper reef habitat, enabling the examination of depth influences on community structure.

@@

Here, we used autonomous reef monitoring structures (ARMS)---standardized collection devices designed to mimic reef structural complexity---to sample cryptic communities while controlling for habitat [@zimmermanArtificialReefMatrix2004; @brainardAutonomousReefMonitoring2009]. Such standardization can reduce variability in recovered communities that may arise due to local variation in habitat composition and availability [@ransomeImportanceStandardizationBiodiversity2017]. We targeted brachyuran crabs, a ubiquitous group often used as an exemplar to assess coral reef biodiversity [@knowltonCoralReefBiodiversity2010; @plaisanceDiversityCoralReefs2011; @servisCharacterizingCoralReef2020; @plaisanceReefassociatedCrustaceanFauna2009] and comprising a notable proportion of the Hawaiian coral reef cryptofauna [284 reported species, @castroCatalogAnomuranBrachyuran2011]. We used assemblages recovered from ARMS units to examine drivers of community composition across the Hawaiian Archipelago. We compared crab communities among islands and regions (MHI vs. NWHI), among major depth zones (shallow vs. mesophotic), and examined the effects of a suite of possible environmental and anthropogenic drivers of community structure on shallow reefs (including fine-scale depth differences), to test hypotheses about their relative importance.

@@

deep_spp <- s %>%
  filter(shallow_deep == "deep") %>%
  arrange(species) %>%
  select(OTU) %>%
  left_join(crab_taxonomy_validated,by=c("OTU" = "otu")) %>%
  pull(display_species) %>%
  unique()
```
Out of the `r all_summary$taxonomy_breakdown$families` families recovered, `r fam['shallow']` were exclusive to shallow ARMS recovered from `r dr(cc$crabs_shallow_untransformed)` m and `r fam['deep']` were exclusive to units $\geq$ 30 m (`r format_list(deep_fams)`; Supplementary [@tbl:tbl_supp_family_depth]). `r str_to_sentence(english(gen['shallow']))` genera were found exclusively shallow and `r gen['deep']` exclusively deep (`r format_list(deep_gens)`; Supplementary [@tbl:tbl_supp_genus_depth]). `r str_to_sentence(english(spp['shallow']))` distinct taxa (including those collapsed to genus or family) were exclusive to shallow ARMS units and `r spp['deep']` were found only deep (`r format_list(deep_spp)`; Supplementary [@tbl:tbl_supp_species_depth]). For a more in-depth taxonomic treatment of the assemblage composition of mesophotic vs. shallow ARMS units, see Hurley et al. [-@hurleyAssessmentShallowMesophotic2016].

@@

all_terms <- english::english(nrow(model_terms))
```
### Environmental drivers {-}
Environmental variables were strongly clustered according to island group (MHI vs NWHI) (Supplementary [@fig:fig_supp_pca]). The best dbRDA model (`r permanova(db_modl,stat="F")`) explained `r pc(RsquareAdj(db_modl)$adj.r.squared,0.01)` (adjusted $\mathrm{R}^2$) of the variability in crab communities across the archipelago ([@fig:fig_dbrda]). `r str_to_sentence(english(nterm))` of the `r str_to_lower(english(nrow(model_terms)))` environmental variables tested were significant and sufficiently uncorrelated. The two drivers contributing most to explained variation were `r drivers$term[1]` (F = `r num(drivers$stat[1],0.01)`) and `r drivers$term[2]` (F = `r num(drivers$stat[2],0.01)`) (Supplementary [@tbl:tbl_supp_rda]). These factors also produced the longest effect vectors in the dbRDA plot ([@fig:fig_dbrda]).

@@

In the CAP analysis of island group and depth zone, ARMS units were correctly reassigned to sample groups `r pc(cap_depth$percent/100,0.1)` of the time. The first two linear discriminant axes explained `r pc(cap_var$x,0.1)` and `r pc(cap_var$y,0.1)` of the variation, respectively. Mesophotic samples (30--90 m) from O&#x02BB;ahu grouped separately from the shallow samples collected across the entire ~2,500 km expanse of the archipelago ([@fig:fig_cap]), whereas shallow samples from O&#x02BB;ahu primarily clustered with shallow samples from other islands ([@fig:fig_cap]). Variability among 30--90 m samples was apparently greater than among all shallow samples, with mean Euclidean distances between deep samples in the CAP ordination significantly greater than between shallow samples (`r t_test(distance_t)`; [@fig:fig_cap], Supplementary [@fig:fig_supp_euclid]). Similarly, variability among samples from O&#x02BB;ahu (including mesophotic depths) exceeded variability among all shallow samples combined (based on Euclidean distance among CAP points, as above; `r t_test(oahu_t,plain=TRUE)`).

@@

ggplot(pcap,aes(x=LD1,y=LD2,fill=sample_group)) +
  stat_ellipse(geom="polygon",level=0.95,alpha=0.1,show.legend = FALSE) +
  # geom_point(aes(shape=oahu),color="black",shape=21,size=5)  +
  geom_point(mapping=aes(shape=oahu),color="black",size=5)  +
  geom_label_repel(data=centroids,aes(label=sample_group), color="white", show.legend = FALSE) +
  scale_shape_manual(values=c("oahu" = 23, "ffs" = 22, "noahu" = 21),name="Island", labels=c("noahu"="Other islands","oahu"="O‘ahu","ffs"="Lalo (French Frigate Shoals)" )) +
  scale_fill_manual(values=sample_depth_pal,name="Sample group") +
  guides(
    fill=guide_legend(override.aes=list(shape=21)),
    shape=guide_legend(override.aes=list(fill="black"))
  ) +
  xlab(str_glue("Linear discriminant 1 [{pc(cap_var$x,0.1)}]")) +
  ylab(str_glue("Linear discriminant 2 [{pc(cap_var$y,0.1)}]")) +
  plotz_theme()

# centroids <- pcap %>% group_by(sample_group) %>%
#   summarise(LD1=mean(LD1),LD2=mean(LD2))
# pcap <- pcap %>%
#   mutate(
#     oahu = factor(
#       if_else(island == "Oahu","oahu","noahu"),
#       levels = c("oahu","noahu")
#     )
#   )

# ggplot(pcap,aes(x=LD1,y=LD2,fill=sample_group)) +
#   stat_ellipse(geom="polygon",level=0.95,alpha=0.1,show.legend = FALSE) +
#   # geom_point(aes(shape=oahu),color="black",shape=21,size=5)  +
#   geom_point(mapping=aes(shape=oahu),color="black",size=5)  +
#   geom_label_repel(data=centroids,aes(label=sample_group), color="white", show.legend = FALSE) +
#   scale_shape_manual(values=c("oahu" = 23, "noahu" = 21),name="Island", labels=c("noahu"="Other islands","oahu"="O‘ahu")) +
#   scale_fill_manual(values=sample_depth_pal,name="Sample group") +
#   guides(
#     fill=guide_legend(override.aes=list(shape=21)),
#     shape=guide_legend(override.aes=list(fill="black"))
#   ) +
#   xlab(str_glue("Linear discriminant 1 [{pc(cap_var$x,0.1)}]")) +
#   ylab(str_glue("Linear discriminant 2 [{pc(cap_var$y,0.1)}]")) +
#   plotz_theme()
cat("{#fig:fig_cap}")
```

@@

While depth was a significant driver of community composition among shallow-water ARMS units ([@fig:fig_dbrda], Supplementary [@tbl:tbl_supp_rda]), its influence was considerably more striking when comparing shallow to mesophotic habitat. In our CAP ordination, all shallow assemblages across the archipelago clustered separately from the deep assemblages $\geq$ 30 m at the single island of O&#x02BB;ahu ([@fig:fig_cap]). Additionally, the two deepest sampled depths (60 m and 90 m) clustered apart from 30 m. According to this analysis, there was more community variation among samples spanning 90 m of depth on a single island than across ~2,500 km of shallow habitat experiencing multiple gradients of environmental, ecological, and anthropogenic influence. This is further supported by the fact that the average Euclidean distance among O&#x02BB;ahu samples (including mesophotic) in canonical space was significantly greater than the average distance among all shallow samples (MHI & NWHI). Shallow crab communities from O&#x02BB;ahu appear to share more affinity with Hōlanikū (~2,200 km distant) than they do with those at 90 m depth on the same reef slope ([@fig:fig_cap]). The CAP also revealed two groups within the mesophotic samples: one at 30 m and the other encompassing the 60--90 m depths. These depth zones correspond with the upper and lower mesophotic zones, respectively [@kahngCommunityEcologyMesophotic2010; @rochaMesophoticCoralEcosystems2018; but see @pyleMesophoticCoralEcosystems2019]. The upper mesophotic zone has been found to constitute an extension of shallow reefs (with beta diversity patterns largely driven by nestedness), whereas the lower mesophotic harbors unique communities (with beta diversity driven primarily by turnover) [@rochaMesophoticCoralEcosystems2018; @slatteryFunctionStabilityMesophotic2024a; @bellwoodRegionalscaleAssemblyRules2001; @kahngCommunityEcologyMesophotic2010]. The evident groupings in the CAP ordination match these trends, with 30 m samples lying closer to shallow O&#x02BB;ahu samples and 60--90 m samples forming a distinct cluster. This pattern of community transition from shallow to mesophotic reefs is consistent with what has been shown for various other taxonomic groups [@lesserGlobalCommunityBreaks2019; @rochaMesophoticCoralEcosystems2018], lending support to the hypothesis that there may be broad-scale depth-based assembly rules for coral reef communities.

@@

A total of 89 ARMS units were deployed over the course of six years (2010--2016) in shallow (`r dr(cc$crabs_shallow_untransformed)` m depth) coral reefs in both the Main (MHI) and Northwest Hawaiian Islands (NWHI) ([@fig:fig_map; @tbl:tbl_arms]). All units included in this study were deployed in forereef habitat. To explore the effects of depth, we also present data from an earlier study by Hurley et al. [-@hurleyAssessmentShallowMesophotic2016], in which 27 units were deployed off O&#x02BB;ahu from 2009--2012 in `r format_range(cc$crab_data %>% sample_tibble() %>% filter(island == "Oahu") %>% pull(depth))` m depth; nine shallow units were deployed at 12 m, and six each at depths of 30 m, 60 m, and 90 m ([@tbl:tbl_arms]). Shallow units were deployed according to NOAA-RAMP methodology, whereas units $\geq$ 30 m were deployed and recovered using technical mixed gas closed-circuit rebreather diving [see @hurleyAssessmentShallowMesophotic2016]. Except when discussing specific depth zones, we will refer to any units deployed at `r dr(cc$crabs_shallow_untransformed)` m as "shallow" and those deployed $\geq$ 30 m as "deep".

@@

ARMS units were retrieved by encapsulating them in-situ in a 100 μm mesh-lined crate to prevent escape of motile organisms. Once returned to the surface, they were placed in containers of aerated, filtered seawater, crates were removed, and units were systematically disassembled plate-by-plate [see @lerayDNABarcodingMetabarcoding2015 for processing details]. All adult brachyuran crabs ($\geq$ 2 mm) were photographed, identified, preserved in 95% ethanol, and retained in a specimen archive at the Hawai&#x02BB;i Institute of Marine Biology. Specimens were initially identified to family or genus using Poore & Ahyong [-@pooreMarineDecapodCrustacea2023] and additionally to species where possible according to expert advice from Scott Godwin (former curator of invertebrate zoology, Bishop Museum) and Gustav Paulay (curator of invertebrate zoology, Florida Museum of Natural History). Where species-level identifications were not possible, taxa were collapsed to family or genus (depending on individual identification). We recognize that the species-level taxonomy of Indo-Pacific brachyurans remains poorly resolved, particularly for abundant groups like the Chlorodiellinae [@lasleyjrIntegrativeSystematicRevision2014; @lasleyjrPhylogeneticRelationshipsUbiquitous2015], and some "species" are known to comprise complexes [e.g. "*Perinia tumida*", @servisCharacterizingCoralReef2020; @plaisanceReefassociatedCrustaceanFauna2009]. However, the prevalence of species complexes and cryptic diversity suggests that our assignments are likely to under- rather than overrepresent the true number of taxa. Thus, any significant patterns we detected likely persist in spite of the potential presence of cryptic diversity. Since our primary concern was categorical diversity of assemblages rather than accurate naming, we have retained species names in cases where complexes are suspected.

@@

We used distance-based redundancy analysis (dbRDA; a form of multivariate multiple regression) with `dbrda` to examine the relative contribution of environmental variables ([@tbl:tbl_env]) to shallow crab community variation across the islands, conditioning the recovery year as a random effect. From a model incorporating all potential drivers, we sequentially removed variables with the largest variance inflation factors (VIF; calculated using `vif.cca`) until all values were <3, indicating variables were sufficiently uncorrelated. Model significance was assessed using permutation tests with `r num(PERM,1)` permutations with the function `anova`. Continuous variables were scaled to unit variance. dbRDA analyses were conducted only for communities sampled from shallow (\< 30 m) ARMS units, since detailed environmental/ecological drivers were not available for deep ($\geq$ 30 m) samples.

@@

Potential drivers of brachyuran crab community structure were assembled from previous studies [@selkoeDNACoralReef2016; @halpernRecentPaceChange2019], NOAA's National Coral Reef Monitoring Program (NCRMP), the Pacific Islands Ocean Observing System (PacIOOS), and MODIS satellite data products compiled by NASA's Physical Oceanography/Ocean Biology Distributed Active Archive Centers [@nasagoddardspaceflightcenterModerateresolutionImagingSpectroradiometer2022; @nasaobpgMODISAquaGlobal2020]. Variables were either intrinsic (site depth, latitude, longitude, island slope) or derived from remote sensing and/or modeling (chlorophyll-A, sea-surface temperature, coral cover, potential larval immigration, cumulative human impacts) ([@tbl:tbl_env], Supplementary [@tbl:tbl_supp_arms_metadata]). Sea-surface temperature (SST) and chlorophyll-A values were calculated by averaging monthly means across the duration of the soaking period for each deployment using AQUA MODIS 4 km-resolution gridded datasets. The cumulative human impacts index, a dimensionless numerical value representing spatially-explicit cumulative anthropogenic impacts on various marine ecosystems, is based on a synthesis of a combination of "high resolution, annual data on the intensity of 14 human stressors and their impact on 21 marine ecosystems over 11 years (2003--2013)" [@halpernGlobalMapHuman2008; @halpernRecentPaceChange2019]. The index of potential larval immigration was imported from [@selkoeDNACoralReef2016] and represents the potential of a given island to act as a destination for larval settlement. It is an "estimate of in-closeness centrality" from a biophysical model simulating larval dispersal in the Hawaiian Islands [@selkoeDNACoralReef2016 supp. mat.]. For details regarding data import, preparation, and synthesis, see Supplementary Information. Due to logistical and dataset limitations, environmental/ecological drivers were only available for shallow (\< 30 m) samples.

@@

@@

(p1 + p2) / (p3 + p4) &
  plot_annotation(tag_levels = list(str_glue("({letters[1:4]})"))) &
  theme(plot.tag = element_text(face="italic"))
cat("{#fig:fig_supp_alpha}")
```

@@

(p1 + p3) / (p2 + p4) &
  plot_annotation(tag_levels = list(str_glue("({letters[1:4]})"))) &
  theme(plot.tag = element_text(face="italic"))
cat("{#fig:fig_supp_alpha_main}")
```

@@

(p1 + p2) &
  plot_annotation(tag_levels = list(str_glue("({letters[1:2]})"))) &
  theme(plot.tag = element_text(face="italic"))
cat("{#fig:fig_supp_alpha2}")
```

@@

```{r fig-supp-pca, fig.cap='Principal component analysis (PCA) of environmental variables by sampling site. Ellipses represent 95% confidence intervals around island groups.', results='asis', dev='svg', fig.width=8,fig.height=8, dependson='init'}
cds <- cc$crab_data_shallow

pcdata <- cds %>%
  select(where(is.numeric),-c(lon,lat),-closest_island) %>%
  select(deployment_year,depth,chl,sst,slope:human_impact,-contains('radius'))

@@

bp <- cds %>%
  bind_cols(scores(pca))

loadings <- pca$rotation %>%
  as_tibble(rownames = "label") %>%
  mutate(
    PC1 = PC1*3,
    PC2 = PC2*3,
    hjust = if_else(PC1 < 0,1,0),
    xnudge = if_else(PC1 < 0,-0.02,0.02)
  )

ggplot(bp,aes(x=PC1,y=PC2)) +
  geom_point(aes(fill=island_group),shape=21,color="black",size=4) +
  stat_ellipse(geom="polygon",aes(group=island_group,fill=island_group),color=NA,alpha=0.1,show.legend = FALSE) +
  # geom_mark_ellipse(color="black",alpha=0.2,show.legend=FALSE, linewidth=0.07) +
  scale_fill_manual(values=island_group_pal,name="Island group") +
  geom_segment(data=loadings,aes(x=0,y=0,xend=PC1,yend=PC2),color="black",size=0.5,arrow=arrow(length=unit(0.02,"npc"))) +
  geom_label(data=loadings,aes(x=PC1+xnudge,y=PC2,label=label,hjust=hjust)) +
  xlab(str_glue("PC1 [{pc(explained[1])}]")) +
  ylab(str_glue("PC2 [{pc(explained[2])}]")) +
  plotz_theme()
cat("{#fig:fig_supp_pca}")
```

@@

@@

```{r tbl-supp-rda,results="asis",dependson="rda-analysis"}
print_table(rda_table,"tbl_supp_rda","Significant marginal terms in the dbRDA analysis of crab community variation across the Hawaiian Archipelago",hdr_case="none",accuracy=0.001)
```

>>>>

The Hawaiian Archipelago, spanning ~2,500 km in the North Pacific Ocean, provides a natural laboratory to examine drivers of metazoan community structure. This chain of islands comprises two regions, the inhabited high islands of the main Hawaiian Islands (MHI) and the uninhabited low islands, atolls, and banks of the Northwestern Hawaiian Islands (NWHI) which reside within one of the world's largest marine protected areas---the Papahānaumokuākea Marine National Monument. The island chain is subject to considerable anthropogenic and physical gradients driving differences in temperature, coral cover, human population, and anthropogenic impact, among others, all of which influence coral reef communities [@selkoeMapHumanImpacts2009]. Nearshore drop-offs along the island chain also allow access to deeper reef habitat, enabling the examination of depth influences on community structure.

@@

Here, we used autonomous reef monitoring structures (ARMS)---standardized collection devices designed to mimic reef structural complexity---to sample cryptic communities while controlling for habitat [@zimmermanArtificialReefMatrix2004; @brainardAutonomousReefMonitoring2009]. Such standardization can reduce variability in recovered communities that may arise due to local variation in habitat composition and availability [@ransomeImportanceStandardizationBiodiversity2017]. We targeted brachyuran crabs, a ubiquitous group often used as an exemplar to assess coral reef biodiversity [@knowltonCoralReefBiodiversity2010; @plaisanceDiversityCoralReefs2011; @servisCharacterizingCoralReef2020; @plaisanceReefassociatedCrustaceanFauna2009] and comprising a notable proportion of the Hawaiian coral reef cryptofauna (284 reported species [@castroCatalogAnomuranBrachyuran2011]). We used assemblages recovered from ARMS units to examine drivers of community composition across the Hawaiian Archipelago. We compared crab communities among islands and regions (MHI vs. NWHI), among major depth zones (shallow vs. mesophotic), and examined the effects of a suite of possible environmental and anthropogenic drivers of community structure on shallow reefs (including fine-scale depth differences), to test hypotheses about their relative importance.

@@

deep_spp <- s %>%
  filter(shallow_deep == "deep") %>%
  arrange(species) %>%
  select(OTU) %>%
  left_join(crab_taxonomy_validated,by=c("OTU" = "otu")) %>%
  pull(display_species) %>%
  unique()
```
Out of the `r all_summary$taxonomy_breakdown$families` families recovered, `r fam['shallow']` were exclusive to shallow ARMS recovered from `r dr(cc$crabs_shallow_untransformed)` m and `r fam['deep']` were exclusive to units $\geq$ 30 m (`r format_list(deep_fams)`; Supplementary [@tbl:tbl_supp_family_depth]). `r str_to_sentence(english(gen['shallow']))` genera were found exclusively shallow and `r gen['deep']` exclusively deep (`r format_list(deep_gens)`; Supplementary [@tbl:tbl_supp_genus_depth]). `r str_to_sentence(english(spp['shallow']))` distinct taxa (including those collapsed to genus or family) were exclusive to shallow ARMS units and `r spp['deep']` were found only deep (`r format_list(deep_spp)`; Supplementary [@tbl:tbl_supp_species_depth]). For a more in-depth taxonomic treatment of the assemblage composition of mesophotic vs. shallow ARMS units, see [@hurleyAssessmentShallowMesophotic2016].

@@

all_terms <- english::english(nrow(model_terms))
```
### Environmental drivers {-}
The PCA of environmental variables indicated strong clustering according to island group (Supplementary [@fig:fig_supp_pca]), with apparent orthogonal variation in variables associated with the MHI vs the NWHI. The best dbRDA model (`r permanova(db_modl,stat="F")`) explained `r pc(RsquareAdj(db_modl)$adj.r.squared,0.01)` (adjusted $\mathrm{R}^2$) of the variability in crab communities across the archipelago ([@fig:fig_dbrda]). `r str_to_sentence(english(nterm))` of the `r str_to_lower(english(nrow(model_terms)))` environmental variables tested were significant and sufficiently uncorrelated. The two drivers contributing most to explained variation were `r drivers$term[1]` (F = `r num(drivers$stat[1],0.01)`) and `r drivers$term[2]` (F = `r num(drivers$stat[2],0.01)`) ([@tbl:tbl_rda]). These factors also produced the longest effect vectors in the dbRDA plot ([@fig:fig_dbrda]).

@@

In the CAP analysis of island group and depth zone, ARMS units were correctly reassigned to sample groups `r pc(cap_depth$percent/100,0.1)` of the time. The first two linear discriminant axes explained `r pc(cap_var$x,0.1)` and `r pc(cap_var$y,0.1)` of the variation, respectively. Mesophotic samples (30--90 m) from O&#x02BB;ahu were grouped separately in ordination space from all shallow samples across the archipelago ([@fig:fig_cap]). Shallow samples from O&#x02BB;ahu primarily clustered with shallow samples from other islands ([@fig:fig_cap]). Variability among 30--90 m samples was apparently greater than among all shallow samples, with mean Euclidean distances between deep samples in canonical space significantly greater than between shallow samples (`r t_test(distance_t)`; [@fig:fig_cap], Supplementary [@fig:fig_supp_euclid]). Similarly, variability among samples from O&#x02BB;ahu (including mesophotic depths) exceeded variability among all shallow samples combined (based on Euclidean distances as above; `r t_test(oahu_t,plain=TRUE)`).

@@

ggplot(pcap,aes(x=LD1,y=LD2,fill=sample_group)) +
  stat_ellipse(geom="polygon",level=0.95,alpha=0.1,show.legend = FALSE) +
  # geom_point(aes(shape=oahu),color="black",shape=21,size=5)  +
  geom_point(mapping=aes(shape=oahu),color="black",size=5)  +
  geom_label_repel(data=centroids,aes(label=sample_group), color="white", show.legend = FALSE) +
  scale_shape_manual(values=c("oahu" = 23, "ffs" = 22, "noahu" = 21),name="Island", labels=c("noahu"="Other islands","oahu"="O‘ahu","ffs"="Lalo (French Frigate Shoals)" )) +
  scale_fill_manual(values=sample_depth_pal,name="Sample grouping") +
  guides(
    fill=guide_legend(override.aes=list(shape=21)),
    shape=guide_legend(override.aes=list(fill="black"))
  ) +
  xlab(str_glue("Linear discriminant 1 [{pc(cap_var$x,0.1)}]")) +
  ylab(str_glue("Linear discriminant 2 [{pc(cap_var$y,0.1)}]")) +
  plotz_theme()

cat("{#fig:fig_cap}")
```

@@

While depth was a significant driver of community composition among shallow-water ARMS units ([@fig:fig_dbrda, @tbl:tbl_rda]), its influence was considerably more striking when comparing shallow to mesophotic habitat. In our CAP ordination, all shallow assemblages across the archipelago clustered separately from the deep assemblages $\geq$ 30 m at the single island of O&#x02BB;ahu ([@fig:fig_cap]). Additionally, the two deepest sampled depths (60 m and 90 m) clustered apart from 30 m. According to this analysis, there was more community variation among samples spanning 90 m of depth on a single island than across ~2,500 km of shallow habitat experiencing multiple gradients of environmental, ecological, and anthropogenic influence. This is further supported by the fact that the average Euclidean distance among O&#x02BB;ahu samples (including mesophotic) in canonical space was significantly greater than the average distance among all shallow samples (MHI & NWHI). Shallow crab communities from O&#x02BB;ahu appear to share more affinity with Hōlanikū (~2,200 km distant) than they do with those at 90 m depth on the same reef slope ([@fig:fig_cap]). The CAP also revealed two groups within the mesophotic samples: one at 30 m and the other encompassing the 60--90 m depths. These depth zones correspond with the upper and lower mesophotic zones, respectively [@kahngCommunityEcologyMesophotic2010; @rochaMesophoticCoralEcosystems2018; but see @pyleMesophoticCoralEcosystems2019]. The upper mesophotic zone has been found to constitute an extension of shallow reefs (with beta diversity patterns largely driven by nestedness), whereas the lower mesophotic harbors unique communities (with beta diversity driven primarily by turnover) [@rochaMesophoticCoralEcosystems2018; @slatteryFunctionStabilityMesophotic2024a; @bellwoodRegionalscaleAssemblyRules2001; @kahngCommunityEcologyMesophotic2010]. The evident groupings in the CAP ordination match these trends, with 30 m samples lying closer to shallow O&#x02BB;ahu samples and 60--90 m samples forming a distinct cluster. This pattern of community transition from shallow to mesophotic reefs is consistent with what has been shown for various other taxonomic groups [@lesserGlobalCommunityBreaks2019; @rochaMesophoticCoralEcosystems2018], lending support to the hypothesis that there may be broad-scale depth-based assembly rules for coral reef communities.

@@

A total of 89 ARMS units were deployed over the course of six years (2010--2016) in shallow (`r dr(cc$crabs_shallow_untransformed)` m depth) coral reefs in both the Main (MHI) and Northwest Hawaiian Islands (NWHI) ([@fig:fig_map; @tbl:tbl_arms]). All units included in this study were deployed in forereef habitat. To explore the effects of depth, we also present data from an earlier study by Hurley et al. [-@hurleyAssessmentShallowMesophotic2016], in which 27 units were deployed off O&#x02BB;ahu from 2009--2012 in `r format_range(cc$crab_data %>% sample_tibble() %>% filter(island == "Oahu") %>% pull(depth))` m depth; nine shallow units were deployed at 12 m, and six each at depths of 30 m, 60 m, and 90 m ([@tbl:tbl_arms]). Shallow units were deployed according to NOAA-RAMP methodology, whereas units $\geq$ 30 m were deployed and recovered using technical mixed gas closed-circuit rebreather diving (see [@hurleyAssessmentShallowMesophotic2016]). Except when discussing specific depth zones, we will refer to any units deployed at `r dr(cc$crabs_shallow_untransformed)` m as "shallow" and those deployed $\geq$ 30 m as "deep".

@@

ARMS units were retrieved by encapsulating them in-situ in a 100 μm mesh-lined crate to prevent escape of motile organisms. Once returned to the surface, they were placed in containers of aerated, filtered seawater, crates were removed, and units were systematically disassembled plate-by-plate (see [@lerayDNABarcodingMetabarcoding2015] for processing details). All adult brachyuran crabs ($\geq$ 2 mm) were photographed, identified, preserved in 95% ethanol, and retained in a specimen archive at the Hawai&#x02BB;i Institute of Marine Biology. Specimens were initially identified to family or genus using Poore & Ahyong [-@pooreMarineDecapodCrustacea2023] and additionally to species where possible according to expert advice from Scott Godwin (former curator of invertebrate zoology, Bishop Museum) and Gustav Paulay (curator of invertebrate zoology, Florida Museum of Natural History). Where species-level identifications were not possible, taxa were collapsed to family or genus (depending on individual identification). We recognize that the species-level taxonomy of Indo-Pacific brachyurans remains poorly resolved, particularly for abundant groups like the Chlorodiellinae [@lasleyjrIntegrativeSystematicRevision2014; @lasleyjrPhylogeneticRelationshipsUbiquitous2015], and some "species" are known to comprise complexes [e.g. "*Perinia tumida*", @servisCharacterizingCoralReef2020; @plaisanceReefassociatedCrustaceanFauna2009]. However, the prevalence of species complexes and cryptic diversity suggests that our assignments are likely to under- rather than overrepresent the true number of taxa. Thus, any significant patterns we detected likely persist in spite of the potential presence of cryptic diversity. Since our primary concern was categorical diversity of assemblages rather than accurate naming, we have retained species names in cases where complexes are suspected.

@@

To visualize how potential environmental drivers varied across the archipelago, we used a principal components analysis (PCA), plotting the results with points colored by island group (MHI or NWHI). We used distance-based redundancy analysis (dbRDA; a form of multivariate multiple regression) with `dbrda` to examine the relative contribution of environmental variables ([@tbl:tbl_env]) to shallow crab community variation across the islands, conditioning the recovery year as a random effect. From a model incorporating all potential drivers, we sequentially removed variables with the largest variance inflation factors (VIF; calculated using `vif.cca`) until all values were <3, indicating variables were sufficiently uncorrelated. Model significance was assessed using permutation tests with `r num(PERM,1)` permutations with the function `anova`. Continuous variables were scaled to unit variance. dbRDA analyses were conducted only for communities sampled from shallow (\< 30 m) ARMS units, since detailed environmental/ecological drivers were not available for deep ($\geq$ 30 m) samples.

@@

We assembled potential drivers of brachyuran crab community structure from previous studies [@selkoeDNACoralReef2016; @halpernRecentPaceChange2019], NOAA's National Coral Reef Monitoring Program (NCRMP), the Pacific Islands Ocean Observing System (PacIOOS), and MODIS satellite data products compiled by NASA's Physical Oceanography/Ocean Biology Distributed Active Archive Centers [@nasagoddardspaceflightcenterModerateresolutionImagingSpectroradiometer2022; @nasaobpgMODISAquaGlobal2020]. Variables were either intrinsic (site depth, latitude, longitude, island slope) or derived from remote sensing and/or modeling (chlorophyll-A, sea-surface temperature, coral cover, potential larval immigration, cumulative human impacts) ([@tbl:tbl_env], Supplementary [@tbl:tbl_supp_arms_metadata]). Sea-surface temperature (SST) and chlorophyll-A values were calculated by averaging monthly means across the duration of the soaking period for each deployment using AQUA MODIS 4 km-resolution gridded datasets. The cumulative human impacts index, a dimensionless numerical value representing spatially-explicit cumulative anthropogenic impacts on various marine ecosystems, is based on a synthesis of a combination of "high resolution, annual data on the intensity of 14 human stressors and their impact on 21 marine ecosystems over 11 years (2003--2013)" [@halpernGlobalMapHuman2008; @halpernRecentPaceChange2019]. The index of potential larval immigration was imported from [@selkoeDNACoralReef2016] and represents the potential of a given island to act as a destination for larval settlement. It is an "estimate of in-closeness centrality" from a biophysical model simulating larval dispersal in the Hawaiian Islands [@selkoeDNACoralReef2016 supp. mat.]. For details regarding data import, preparation, and synthesis, see Supplementary Information. Due to logistical and dataset limitations, environmental/ecological drivers were only available for shallow (\< 30 m) samples.

@@

# Tables {-}
```{r tbl-rda,results="asis",dependson="rda-analysis"}
print_table(rda_table,"tbl_rda","Significant marginal terms in the dbRDA analysis of crab community variation across the Hawaiian Archipelago",hdr_case="none",accuracy=0.001)
```

@@

(p1 + p2) / (p3 + p4) &
  plot_annotation(tag_levels = list(letters[1:4])) &
  theme(plot.tag = element_text(face="bold"))
cat("{#fig:fig_supp_alpha}")
```

@@

(p1 + p3) / (p2 + p4) &
  plot_annotation(tag_levels = list(letters[1:4])) &
  theme(plot.tag = element_text(face="bold"))
cat("{#fig:fig_supp_alpha_main}")
```

@@

(p1 + p2) &
  plot_annotation(tag_levels = list(letters[1:2])) &
  theme(plot.tag = element_text(face="bold"))
cat("{#fig:fig_supp_alpha2}")
```

@@

```{r fig-supp-pca, fig.cap='Principal component analysis (PCA) of environmental variables by sampling site. Ellipses show 95% confidence intervals around island groups.', results='asis', dev='svg', fig.width=8,fig.height=8, dependson='init'}
cds <- cc$crab_data_shallow

termz <- model_terms %>%
  select(variable,display=`Environmental Driver`)

pcdata <- cds %>%
  select(where(is.numeric),-closest_island) %>%
  select(recovery_year,lon,lat,depth,chl,sst,slope:human_impact,-contains('radius'))

@@

bp <- cds %>%
  bind_cols(scores(pca)) %>%
	mutate(island_group = str_to_sentence(island_group))

loadings <- pca$rotation %>%
  as_tibble(rownames = "label") %>%
  mutate(
    PC1 = PC1*3,
    PC2 = PC2*3,
    hjust = if_else(PC1 < 0,1,0),
    xnudge = if_else(PC1 < 0,-0.02,0.02)
  ) %>%
  left_join(termz,by=c(label="variable")) %>%
  mutate(
    label = str_to_sentence( str_replace_all(label,"_"," ") ),
    label = coalesce(display,label)
  ) %>%
  select(-display)

ggplot(bp,aes(x=PC1,y=PC2)) +
  geom_point(aes(fill=island_group),shape=21,color="black",size=4) +
  stat_ellipse(geom="polygon",aes(group=island_group,fill=island_group),color=NA,alpha=0.1,show.legend = FALSE) +
  scale_fill_manual(values=island_group_pal,name="Island group") +
  geom_segment(data=loadings,aes(x=0,y=0,xend=PC1,yend=PC2),color="black",size=0.5,arrow=arrow(length=unit(0.02,"npc"))) +
  geom_label(data=loadings,aes(x=PC1+xnudge,y=PC2,label=label,hjust=hjust)) +
  xlab(str_glue("PC1 [{pc(explained[1])}]")) +
  ylab(str_glue("PC2 [{pc(explained[2])}]")) +
  plotz_theme()
cat("{#fig:fig_supp_pca}")
```

@@

```{r fig-supp-euclid, fig.cap='Box plot of Euclidean distances between points in canonical space based on CAP analysis of crab assemblages by depth zone. Asterisk indicates significant difference (Welch two-sample t-test).', results="asis", dev="svg", fig.width=6, fig.height=4, dependson="cap-analysis"}
dists <- bind_rows(
  tibble(dist=shallow_dist,depth="Shallow"),
  tibble(dist=deep_dist,depth="Deep")
) %>%
  mutate(depth = factor(depth,levels=c("Shallow","Deep")))

pointsize <- 5
ggplot(dists,aes(x=depth,y=dist,fill=depth)) +
  geom_boxplot(notch=TRUE) +
  geom_point(aes(x,y),data=sig,shape=8,size=pointsize,stroke=1.5,fill="black") +
  scale_fill_manual(values=sample_depth_pal) +
  guides(fill="none") +
  plotz_theme() +
  xlab("Sampling depth") +
  ylab("Euclidean distance")
cat("{#fig:fig_supp_euclid}")
```

@@
